Forecasting Gamma-ray Bursts using Gravitational Waves

### 2018-03-30 Paper/Manuscript.tex
Updates during Cotswold weekend

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars early enough to alert the electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very start.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the inteferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc in the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc in the Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time thus we expect no forecast of GRBs in the 2020s. 
On the other hand, we demonstrate that Einstein Telescope will provide us with warning times as long as a few hours.  
Taking one hour as a benchmark warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS event rates within this horizon distance, we show that we expect to forecast $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
%Our main conclusions are: (i) we expect ALV network to forecast 1 to 3 GRBs in the 2020s, (ii)

@@

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced Ligo-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory (\cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $\sim 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched the largest-ever campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
more than half a year after the initial GRB \cite{GBM:2017lvd}.

@@

A week after GW170817, the ALV network was taken offline for upgrades for the third observation run (O3) 
scheduled to start in Autumn 2018 with roughly twice the sensitivity of O2.
In addition, the Japanese cryogenic interferometer KAGRA (\cite{KAGRA, KAGRA2}) will start its test runs in 2018-2019 and reach design sensitivity circa 2021-2022 \cite{Akutsu:2017thy, Aasi:2013wya}. By 2025, LIGO-India detector will join the global interferometer network hence improving both the overall sensitivity and the sky-localization capabilities \cite{Aasi:2013wya}.
Around this time, the construction of the first third-generation ground-based interferometer, Einstein Telescope, will begin in Europe \cite{ET_doc}. 
Meanwhile, the US is currently considering a mid-2020s update to LIGO called Voyager \cite{LIGO_Voy} and the construction of a more ambitious 40\,km long interferometer dubbed The Cosmic Explorer \cite{CE}.
All of these near-future improvements to the GW-detector network beg a simple question: can we detect the inspiral of a binary neutron star system
early enough to witness a gamma-ray burst as it happens? In other words, can we forecast GRBs using GW detections?

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning the GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time difference
between the merger and the instant of detection (detector trigger) defined in terms of a certain threshold signal-to-noise ratio.
To this end, we consider the $\sim 2020$ ALV network composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

@@

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by radiation reaction of GW emission from the system due to the time variance of the quadrupole moment of the binary.
We nonetheless supplement our Newtonian (leading-order) time evolution with results from post-Newtonian theory up to 3.5 post-Newtonian order \cite{Blanchet_LRR}.
The evolution of a BNS system in the strong-field regime, where speeds $\gtrsim 0.1c$, is a very complex and complicated problem requiring fully relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its simplicity, our leading-order treatment is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

@@

This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding GW data.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020}, \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we reapply the above procedure to black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

@@

\section{Binary systems composed of neutron stars}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution of BNS}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
This is the celebrated Einstein quadrupole formula \cite{1918SPAW}. 
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power generated %in gravitational radiation
by a system with changing mass moments.

@@

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}, \label{eq:Edot2}
\ee
%
where, recall $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
The quantity in parantheses in Eq.~(\ref{eq:Edot2}) is ubiquitous in post-Newtonian theory and is
referred to as the dimensionless inverse separation $x$.
As $x\sim v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from in Eq.~(\ref{eq:Edot2}) that $\dot{E}\sim x^5$. 
%hence the common statement in the literature saying that gravitational radiation starts

The energy radiated away in GWs decreases the total energy of the binary thus making the orbit more bound. 
Since the total energy is given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,  \label{eq:binding_energy1}
\ee
%
we have $\dot{E}_b \propto \dot{r}/r$; hence $\dot{E}_b < 0$ implies $\dot{r}<0$.
Using Kepler's third law, i.e., $ \Omega^2 = G M r^{-3}$ and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
We can now define an inspiral time at a given GW frequency
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $\simeq 1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$.

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a much larger time scale than $T_K$ itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular (ISCO) orbit in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim 1000\,$Hz for a system with $M\approx 2 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations will be $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \, . \label{eq:Ncyc2}
\ee
%
Restoring some familiar numbers in this expression yields
%
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

Another useful relation is how the inspiral time scales with respect to the orbital radius corresponding to the observed GW frequency.
From Kepler's third law, we immediately have $\dot{r}/r= -2\dot{\Omega}/(3\Omega)=-2\dot{f}/(3f)$, which, via Eq.~(\ref{eq:tau_insp}),
yields $\dot{r}/r= -1/(4\tau)$ which integrates to
%
\be
r(\tau)=r_i \left(\f{\tau}{\tau_i}\right)^{1/4}\label{eq:r_of_tau},
\ee
%
where $r_i, \tau_i=t_\text{coal}-t_i$ are the initial radius and time that the BNS
is ``picked up'' by a detector. Solving $\tau(f_i)=\tau_i$ for $f_i=\Omega_i/\pi$ using Eq.~(\ref{eq:tau_insp}) and rewriting $\Omega_i$ as a function of $r_i$ via
Kepler's third law gives us
%
\be
\tau_i = \f{5}{256}\, \f{c^5 r_i^4}{G^3 M^2 \mu}\simeq 20.37\,\text{hours}\, \left(\f{T_K}{1\,\text{s}}\right)^{8/3} \left(\f{2.8M_\odot}{M}\right)^{2/3}\,\left(\f{0.7M_\odot}{\mu}\right) \label{eq:tau_of_r}.
\ee
%

Let us now turn our attention to the GWs generated by the inspiral.
These are actually tensorial perturbations of flat (Minkowski) spacetime propagating at the speed of light.
In the so-called wave zone, defined by the condition $ c/f \ll D$, where $D$ is the distance to the GW source, 
these tensor modes satisfy the flat spacetime sourceless wave equation $\Box\bar{h}_{\mu\nu} = 0$.
%meaning that we are considering the solutions far from the GW source, 
In a suitable gauge, such as the commonly used transverse-traceless gauge, 
it can be shown that there exist only two physical degrees of freedom which manifest themselves as two independent polarization amplitudes $h_+$ and $h_\times$. %when interacting with masses away from the source.
For a binary in a circular orbit, these states read
%
\begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi_\text{N}(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi_\text{N}(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\be
h_c(t) =\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \label{eq:h_c}
\ee
is the characteristic strain.
$\iota=\cos^{-1}(\mathbf{\hat{n}}\cdot{\mathbf{\hat{L}}}) $ is the inclination angle between the line of sight unit vector $\mathbf{\hat{n}} $ and the orbital angular momentum unit vector $\mathbf{\hat{L}}$. 
For an orbit seen edge on, $\iota = \pi/2$ giving $h_\times=0$, i.e,  a purely plus-polarized wave.
When $\iota=0$ the orbit is seen face-on and we have a circularly-polarized wave: $\langle h_+ \rangle = \langle h_\times\rangle$ where $\langle \ldots \rangle$
denote time-averages over one orbit. Note that as $f$ increases in time, so does $h(t)$ hence the characteristic ``chirping'' of GW signals.

$\Phi_\text{N}(t)$ in Eqs.~(\ref{eq:hplus_TD}, \ref{eq:hcross_TD}) is the phase of the GWs given by
%
\begin{align}
\Phi_\text{N}(t) &=\int_{t_i}^t dt' \omega(t') \nn\\
&= -2\left(\f{5 G M_c}{c^3}\right)^{-5/8} (t_\text{coal}-t)^{5/8}+ \Phi_0 \, .\label{eq:Phase}
\end{align}
%
$\Phi_0 \equiv \Phi(t=t_\text{coal})$ is an integration constant. The subscript N denotes the Newtonian (leading-order) contribution. 
Higher-order contributions can be added in terms a post-Newtonian series which make up $\lesssim 2\%$ of the total phase.
Although we will include contributions up to and including 3.5PN to obtain the results of Sec.~\ref{sec:results}, 
we do not explicitly show these expressions, but they can be found in Ref.~\cite{Blanchet_LRR}.

>>>>

\begin{abstract}
We explore the intriguing possibility of employing future ground-based gravitational-wave interferometers to detect the inspiral of binary neutron stars early enough to alert electromagnetic observatories so that a gamma-ray burst (GRB) can be observed in its entirety from its very start.
We quantify the ability to predict a GRB by computing the time a binary neutron star (BNS) system takes to inspiral from its moment of detection to its final merger. We define the moment of detection to be the instant at which the inteferometer network accumulates a signal-to-noise ratio of 15. %for the BNS inspiral.
For our computations, %of advance warning times, 
we specifically consider BNS systems at luminosity distances of (i) $D\le200\,$Mpc in the three-interferometer Advanced-LIGO-Virgo network of 2020, and (ii) $D \le 1000\,$Mpc in the Einstein Telescope's B and C configurations. 
In the case of Advanced LIGO-Virgo we find that we may at best get a few minutes of warning time thus we expect no forecast of GRBs in the 2020s. 
On the other hand, Einstein Telescope will provide us with advance warning times up to five hours.  
Taking one hour as a benchmark warning time, we obtain a corresponding horizon
distance of roughly 600 Mpc for the Einstein Telescope C configuration.
%, more specifically, greater than an hour
%for BNS inspirals closer than roughly 600\,Mpc. Taking this luminosity distance to be our horizon distance, 
Using current BNS event rates within this horizon distance, we show that this translates to forecasts of $\gtrsim \ord(10^2)$ GRBs in the 2030s. %with Einstein C. %with the C configuration of Einstein Telescope.
We reapply our warning-time computation to binary black hole - neutron star inspirals and find that we expect 1 to 3 tidal disruption events 
to be forecast by the same detector.
%Our main conclusions are: (i) we expect ALV network to forecast 1 to 3 GRBs in the 2020s, (ii)

@@

\section{Introduction}\label{Sec:Intro}
The 6.95\footnote{There have thus far been six detections with $5\sigma$ statistical significance and one additional event, LVT151012 \cite{TheLIGOScientific:2016pea}, with $2\sigma \simeq 95\%$ significance hence 6.95 overall.} 
gravitational-wave merger events detected by the Advanced Ligo-Virgo network have firmly established gravitational-wave astronomy
as an observational science \cite{TheLIGOScientific:2016pea, Abbott:2017gyy, Abbott:2017oio, Abbott:2017vtc, GW170817}. 
Though the first event of 14 September 2015 (GW150914 \cite{GW150914}) will always be the ``poster-child'' of this field,
the icing on the cake was the 17 August 2017 (GW170817 \cite{GW170817}) event involving the inspiral and merger of a binary neutron star system 
promptly followed, 1.7 seconds later, by a gamma-ray burst (GRB170817A) detected by the Fermi Gamma-ray Burst Monitor \cite{Fermi} and by the International Gamma-Ray Astrophysics Laboratory \cite{Svinkin, Savchenko:2017ffs}.
The Advanced-LIGO-Virgo network's initial source localization to within $\sim 31\,\text{degree}^2$ in the Southern skies
enabled astronomers to locate the electromagnetic counterpart in the galaxy NGC 4993, 
first picked up by the One-Meter, Two Hemisphere team less than 11 hours after the merger \cite{Coulter2017, Coulter:2017wya}.
This subsequently launched the largest-ever campaign of multi-messenger astronomy across the entire electromagnetic (EM) spectrum which is still ongoing
more than half a year after the initial GRB \cite{GBM:2017lvd}.

@@

A week after GW170817, the ALV network was taken offline for upgrades for the third observation run (O3) 
scheduled to start in Autumn 2018 with roughly twice the sensitivity of O2.
In addition, the Japanese cryogenic interferometer KAGRA (\cite{KAGRA, KAGRA2}) will start its test runs in 2018-2019 and reach design sensitivity circa 2021-2022 \cite{Akutsu:2017thy, Aasi:2013wya}. By 2025, LIGO-India detector will join the global interferometer network hence improving both the overall sensitivity and the sky-localization capabilities of the interferometer network \cite{Aasi:2013wya}.
Around this time, the construction of the first third-generation ground-based interferometer, Einstein Telescope, will begin in Europe \cite{ET_doc}. 
Meanwhile, the US is currently considering a mid-2020s update to LIGO called Voyager \cite{LIGO_Voy} and the construction of a more ambitious 40\,km long interferometer dubbed The Cosmic Explorer \cite{CE}.
All of these near-future improvements to the GW-detector network beg a simple question: can we detect the inspiral of a binary neutron star system
early enough to witness a gamma-ray burst as it happens? In other words, can we forecast GRBs using GW detections?

@@

Here, we are interested  in exploring the predictive power of near-future ground-based GW detectors. 
%thus leave the task of reducing the latency to the experts.
More specifically, we wish to find out how much early warning the GW detectors can provide us before the merger/GRB.
Ref.~\cite{Cannon:2011vi} approached this from an algorithmic perspective by developing computationally inexpensive methods to detect inspiral
signals in GW data to provide early-warning triggers \emph{before} the merger (also see Ref.~\cite{Knowles:2018hqq} for a very recent improvement to what is currently being
used in the ALV pipeline).
We instead focus on the forecasting capabilities of the ground-based interferometers at their \emph{full design} sensitivity by computing the time difference
between the merger and the instant of detection (detector trigger) defined in terms of a certain threshold signal-to-noise ratio.
To this end, we consider the ALV network of 2020s composed of three L-shaped interferometers operating at their design sensitivities
and Einstein Telescope of 2030s made up of triangular interferometers.
Ref.~\cite{Nissanke:2012dj} investigated the capabilities of three to five-interferometer detector networks of 2020s in terms of their horizon distances
to inspiralling BNSs and their sky-localization of these sources. Our work here is somewhat complementary in the sense that we are mostly concerned with
how early the GW networks can detect the inspiral before the merger.

@@

Our computation for $T_\text{AW}$ is based on the Newtonian evolution of a binary system composed of two point masses in a quasi-circular Keplerian orbit. 
The long-term dynamics is governed by the radiation reaction of 
the GW emission from the system due to the time variance of the quadrupole moment of the binary.
Though the Newtonian (leading-order) evolution is better than $ 95\%$ accurate
we nonetheless supplement it with results from post-Newtonian theory up to 3.5 post-Newtonian order \cite{Blanchet_LRR}.
The evolution of a BNS system in the strong-field regime, where speeds $\gtrsim 0.1c$, is a very complex and complicated problem requiring fully relativistic, magnetohydrodynamic codes 
with neutrino transport running on large-scale computing clusters and taking up to millions of CPU hours 
(see Ref.~\cite{Faber:2012rw} for a review, and Refs.~\cite{Kyutoku:2017voj, Zappa:2017xba, Dietrich:2018upm} for the latest developments and references therein).
Despite its simplicity, our quadrupole radiation driven adiabatic
evolution is more than sufficient to provide reliable estimations of advance warning times.
We support this claim in Sec.~\ref{sec:idealizations} with an extensive list of computations showing how much each of our approximations affects the inspiral time.

@@

This article is organized as follows. Sec.~\ref{sec:BNS_inspiral} introduces the formulation for the leading-order evolution of the binary inspiral.
Sec.~\ref{sec:IFO_response} details the interferometer response to GWs based on topology.
Sec.~\ref{sec:idealizations} lists the various idealizations we employ to simplify our treatment and supplies justification for each.
Sec.~\ref{sec:GW170817} tests our model network and evolution using the parameters of GW170817 and the corresponding GW data.
Sec.~\ref{sec:results} contains our main results presented as advance warning times in Tables~\ref{table:LIGO2020}, \ref{table:ET}; 
and as horizon distances and event rates in Table~\ref{table:horizon}.
In Sec.~\ref{sec:BH_NS}, we recompute the advance warning times in the case of black hole - neutron star inspirals to see whether or not 
future detectors can forecast a tidal disruption event, for which we summarize our findings in Table~\ref{table:ET_BH_NS}.

@@

\section{Binary neutron star inspirals}\label{sec:BNS_inspiral}
\subsection{Newtonian evolution}
For us, the starting point for the evolution of a binary under gravitational radiation reaction
%the emission of GWs 
is the expression for the power emitted in GWs given by 
%for the generated by the time-variance of the mass quadrupole moment
%
\be
\dot{E} = \f{G}{5c^5}\langle \dddot{Q}_{ij}\dddot{Q}_{ij} \rangle \label{eq:Edot_quadrupole}\, .
\ee
%
This is the celebrated Einstein quadrupole formula \cite{1918SPAW}. 
Here, $Q_{ij}$ is a trace-reversed mass quadrupole moment which
is the dominant contribution to the power emitted %in gravitational radiation
by a system with changing mass moments.

@@

An important quantity in GW astronomy is the chirp mass of the binary given by
%
\be
M_c = \mu^{3/5} M^{2/5} = \f{(m_1 m_2)^{3/5}}{(m_1+m_2)^{1/5}} \label{eq:chirp_mass}\, ,
\ee
%
which turns Eq.~(\ref{eq:Edot}) into
%
\be
\dot{E} = \f{32}{5}\f{c^5}{G} \left(\f{G M_c\, \omega}{2c^3}\right)^{10/3}, \label{eq:Edot2}
\ee
%
where $\omega=2\Omega$ is the angular frequency of the quadrupolar GWs. 
The quantity in parantheses in Eq.~(\ref{eq:Edot2}) is ubiquitous in post-Newtonian theory and is
referred to as the dimensionless inverse separation $x$.
As $x\sim v^2/c^2$, it is used to keep track of post-Newtonian orders and we can see from in Eq.~(\ref{eq:Edot2}) that $\dot{E}\sim x^5$. 
%hence the common statement in the literature saying that gravitational radiation starts

The emission of power in GWs decreases the total energy of the binary thus making the orbit more bound. 
The total (binding) energy of the binary is given by
%
\be
E_b = -\frac{G m_1 m_2}{2r}\,  \label{eq:binding_energy1}
\ee
%
giving $\dot{E}_b \propto \dot{r}/r$; hence $\dot{E}_b < 0$ implies $\dot{r}<0$, i.e, the orbital separation decreases in time.
Using Kepler's third law, $ \Omega^2 = G M r^{-3}$, and Eq.~(\ref{eq:chirp_mass}) we obtain
%
\be
E_b = -\left(\f{G^2 M_c^5 \omega^2}{32}\right)^{1/3}\, . \label{eq:binding_energy2}
\ee
%
Energy conservation dictates that $\dot{E}=-\dot{E}_b$. Using the chain rule to determine $\dot{E}_b$ from Eq.~(\ref{eq:binding_energy2})
and setting the resulting expression equal to Eq.~(\ref{eq:Edot2}) yields $\dot{\omega}$ as a function of $\omega$ and various constants.
Translating this expression using $\omega=2\pi f$ we arrive at the standard expression for the GW frequency evolution
%%
%
\be
\dot{f} = \f{96}{5}\pi^{8/3} \f{(G M_c)}{c^5}^{5/3}\, f^{11/3} \label{eq:fdot}
\ee
%
This can be integrated straightforwardly after defining a new time variable $\tau\equiv t_\text{coal}-t$ that equals zero when the neutron stars coalesce.
We can now define an inspiral time at a given GW frequency
%
\be
\tau_\text{insp}(f) = \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}} \,f^{-8/3}\label{eq:tau_insp}\, ,
\ee
%
%
%\be
%\tau_\text{insp}(f) \equiv \f{5}{256\pi}\f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-8/3}-f_\text{ISCO}^{-8/3}\right]\label{eq:tau_insp} \, .
%\ee
%
which we can rewrite as follows
%
\be
\tau_\text{insp}(f) \simeq 16.72\,\text{minutes} \, \left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\,\left(\f{10\,\text{Hz}}{f}\right)^{8/3}, \label{eq:tau_insp2}
\ee
%
where $1.219 M_\odot$ is the chirp mass corresponding to $m_1=m_2=1.4 M_\odot$.

We can also compute the number of GW cycles over the course of an inspiral. Given that the orbital period $T_K(t)$ varies over a time scale much larger than $T_K$ itself
we can write
%
\be
\mathcal{N}_\text{cyc}(f) = \int_f^{f_\text{max}} \f{f'}{\dot{f'}}\, df' =\f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\left[ f^{-5/3}-f_\text{max}^{-5/3}\right] \label{eq:Ncyc1}
\ee
%
We now introduce a cut-off for the inspiral imposed by the frequency of the innermost stable circular (ISCO) orbit in Schwarzschild spacetime
\be
\Omega_\text{ISCO} = \f{c^3}{6^{3/2} G M} \label{eq:Sch_f_isco}.
\ee
%
%
We will see in Sec.~\ref{sec:idealizations} that $f_\text{ISCO} \sim 1000\,$Hz for a system with $M\approx 2 M_\odot$. 
On the other hand, the frequencies of interest for our advance warning time estimations will be $\lesssim 10\,$Hz
meaning that $f^{-5/3} \gg f_\text{ISCO}^{-5/3}$. Therefore, at the leading (Newtonian) order, the number of GW cycles can be approximated by
%
\be
\mathcal{N}_\text{cyc}(f) \approx \f{1}{32\pi} \f{c^5}{(\pi G M_c)^{5/3}}\, f^{-5/3} \,  \label{eq:Ncyc2}
\ee
%
%Restoring some familiar numbers in this expression yields
%
which can be rewritten as
\be
\mathcal{N}_\text{cyc}(f)\approx 1.605\times 10^4 \left(\f{10\,\text{Hz}}{f} \right)^{5/3}\,\left(\f{1.219 M_\odot}{M_c}\right)^{5/3}\label{eq:Ncyc3}\, .
\ee
%
This expression is quite telling: if a ground-based interferometer picks up an
inspiralling BNS at $f=10\,$Hz then there will be $\gtrsim \ord(10^4)$ GW cycles in the detector's data stream until the merger.

Another useful relation is how the inspiral time scales with respect to the orbital radius corresponding to the observed GW frequency.
From Kepler's third law, we immediately have $\dot{r}/r= -2\dot{\Omega}/(3\Omega)=-2\dot{f}/(3f)$, which, via Eq.~(\ref{eq:tau_insp}),
yields $\dot{r}/r= -1/(4\tau)$ which integrates to
%
\be
r(\tau)=r_i \left(\f{\tau}{\tau_i}\right)^{1/4}\label{eq:r_of_tau},
\ee
%
where $r_i, \tau_i=t_\text{coal}-t_i$ are the initial radius and time that the BNS
is ``picked up'' by a detector. Solving $\tau(f_i)=\tau_i$ for $f_i=\Omega_i/\pi$ using Eq.~(\ref{eq:tau_insp}) and rewriting $\Omega_i$ as a function of $r_i$ via
Kepler's third law gives us
%
\be
\tau_i = \f{5}{256}\, \f{c^5 r_i^4}{G^3 M^2 \mu}\simeq 20.37\,\text{hours}\, \left(\f{T_K}{1\,\text{s}}\right)^{8/3} \left(\f{2.8M_\odot}{M}\right)^{2/3}\,\left(\f{0.7M_\odot}{\mu}\right) \label{eq:tau_of_r},
\ee
%
where the seemingly arbitrary $0.7M_\odot$ is the value of $\mu$ for $m_1=m_2=1.4M_\odot$.

Let us now turn our attention to the GWs generated by the inspiral.
These are tensorial perturbations propagating at the speed of light in flat (Minkowski) spacetime.
In the so-called wave zone, defined by the condition $ c/f \ll D$, where $D$ is the distance to the GW source, 
these tensor modes satisfy the flat spacetime sourceless wave equation. %$\Box\bar{h}_{\mu\nu} = 0$.
%meaning that we are considering the solutions far from the GW source, 
In a suitable gauge, such as the commonly used transverse-traceless gauge, 
it can be shown that there exist only two physical degrees of freedom which manifest themselves as two independent polarization amplitudes $h_+$ and $h_\times$. %when interacting with masses away from the source.
For a binary in a circular orbit, these states read
%
\begin{align}
 h_+(t) &= h_c(t)\, \left(\f{1+\cos^2\iota}{2}\right)\, \cos[\Phi_\text{N}(t)],\label{eq:hplus_TD}\\
 h_\times(t) & =h_c(t)\,\cos\iota \sin[\Phi_\text{N}(t)]\label{eq:hcross_TD},
\end{align}
%
%
where
%
\be
h_c(t) =\f{4}{D}\left(\f{G M_c}{c^2}\right)^{5/3}\, \left(\f{\pi f(t)}{c}\right)^{2/3} \label{eq:h_c}
\ee
is the characteristic strain and
$\iota=\cos^{-1}(\mathbf{\hat{n}}\cdot{\mathbf{\hat{L}}}) $ is the inclination angle between the line of sight unit vector $\mathbf{\hat{n}} $ and the orbital angular momentum unit vector $\mathbf{\hat{L}}$. 
For an orbit seen edge on, $\iota = \pi/2$ giving $h_\times=0$, i.e,  a purely plus-polarized wave.
When $\iota=0$ the orbit is seen face-on and we have a circularly-polarized wave: $\langle h_+ \rangle = \langle h_\times\rangle$ where $\langle \ldots \rangle$
denote time-averages over one orbit. Note that as $f$ increases in time, so does $h(t)$ hence the characteristic ``chirping'' of GW signals.

$\Phi_\text{N}(t)$ in Eqs.~(\ref{eq:hplus_TD}, \ref{eq:hcross_TD}) is the phase of the GWs given by
%
\begin{align}
\Phi_\text{N}(t) &=\int_{t_i}^t dt' \omega(t') \nn\\
&= -2\left(\f{5 G M_c}{c^3}\right)^{-5/8} (t_\text{coal}-t)^{5/8}+ \Phi_0 \, ,\label{eq:Phase}
\end{align}
%
where $\Phi_0 \equiv \Phi(t=t_\text{coal})$ is an integration constant. The subscript N denotes the Newtonian (leading-order) contribution. 
Higher-order contributions can be added in terms a post-Newtonian series, which make up $\lesssim 2\%$ of the total phase.
Although we will include contributions up to and including 3.5PN to obtain the results of Sec.~\ref{sec:results}, 
we will not show these expressions here, but they can be found in Ref.~\cite{Blanchet_LRR}.
